---
title: "Getting Started with centrimpact"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with centrimpact}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

The `centrimpact` package provides tools for analyzing and visualizing community-engaged research metrics based on the CEnTR*IMPACT framework (Price, 2024). This framework quantifies four critical dimensions of community-engaged research that go beyond traditional academic metrics:

1. **Alignment** - Shared vision between researchers and partners
2. **Cascade Effects** - Ripple effects across social networks
3. **Dynamics** - Quality of partnership processes
4. **Indicators** - Traditional academic productivity markers

This vignette demonstrates the basic workflow for analyzing each dimension and creating publication-ready visualizations.

## Installation

```{r eval = FALSE}
# Install from GitHub
devtools::install_github("CENTR-IMPACT/centrimpact-review")
```

```{r setup}
library(centrimpact)
```

## Analyzing Project Alignment

The Alignment Score (Sa) quantifies consensus between researchers and community partners across four key areas: Goals, Values, Roles, and Resources. Higher scores indicate stronger shared vision.

### Generate and Analyze Data

```{r alignment-analysis}
# Generate example alignment data
alignment_data <- generate_alignment_data(seed = 36)

# View the structure
head(alignment_data)

# Analyze alignment
alignment_results <- analyze_alignment(alignment_data)

# View results
print(alignment_results)
```

### Visualize with Slopegraph

The slopegraph shows how researcher and partner ratings compare across domains:

```{r alignment-slopegraph, fig.width = 7, fig.height = 5}
# Create slopegraph visualization
plot_slopegraph <- visualize_alignment(alignment_results)
print(plot_slopegraph)
```

### Visualize with Abacus Plot

The abacus plot provides an alternative view of alignment patterns:

```{r alignment-abacus, fig.width = 7, fig.height = 5}
# Create abacus plot
plot_abacus <- visualize_abacus(alignment_results)
print(plot_abacus)
```

## Analyzing Cascade Effects

The Cascade Effects Score (Sc) quantifies how information and power distribute across three degrees of separation from core participants, based on social network analysis principles.

### Generate and Analyze Data

```{r cascade-analysis}
# Generate example cascade data (returns a one-row survey parameter frame)
cascade_data <- generate_cascade_data(seed = 36)

# View the structure
print(cascade_data)

# Analyze cascade effects
cascade_results <- analyze_cascade(cascade_data)

# View results
print(cascade_results)
```

### Visualize with Racetrack Plot

The radial bar chart ("racetrack plot") shows distribution across network degrees:

```{r cascade-plot, fig.width = 7, fig.height = 7}
# Create cascade visualization
plot_cascade <- visualize_cascade(cascade_results)
print(plot_cascade)
```

## Analyzing Project Dynamics

The Project Dynamics Score (Sd) quantifies how well a project follows Community-Based Participatory Research (CBPR) principles (Wallerstein & Duran, 2010; Wallerstein et al., 2020).

### Generate and Analyze Data

```{r dynamics-analysis}
# Generate example dynamics data
dynamics_data <- generate_dynamics_data(seed = 36)

# View the structure
head(dynamics_data)

# Analyze project dynamics
dynamics_results <- analyze_dynamics(dynamics_data)

# View results
print(dynamics_results)
```

### Visualize with Rose Chart

The rose chart displays dynamics across CBPR dimensions:

```{r dynamics-plot, fig.width = 7, fig.height = 7}
# Create dynamics visualization
plot_dynamics <- visualize_dynamics(dynamics_results)
print(plot_dynamics)
```

## Visualizing Project Indicators

Project Indicators capture traditional academic metrics. These require no separate analysis function as they represent direct counts.

### Generate and Visualize Data

```{r indicators-plot, fig.width = 7, fig.height = 5}
# Generate example indicators data
indicators_data <- generate_indicators_data(seed = 36)

# View the structure
head(indicators_data)

# Create horizontal bubble chart
plot_indicators <- visualize_indicators(indicators_data)
print(plot_indicators)
```

## Customizing Visualizations

All visualization functions return `ggplot2` objects, allowing for further customization:

```{r customize-example, fig.width = 7, fig.height = 5}
library(ggplot2)

# Customize alignment plot
plot_slopegraph +
  labs(title = "Researcher-Partner Alignment",
       subtitle = "Community Health Equity Project") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14))
```

## Complete Workflow Example

Here's a complete workflow analyzing all four dimensions for a single project:

```{r complete-workflow, fig.width = 7, fig.height = 5}
# 1. Alignment
alignment_data <- generate_alignment_data()
alignment_results <- analyze_alignment(alignment_data)
alignment_plot <- visualize_alignment(alignment_results)

# 2. Cascade Effects
cascade_data <- generate_cascade_data()
cascade_results <- analyze_cascade(cascade_data)
cascade_plot <- visualize_cascade(cascade_results)

# 3. Dynamics
dynamics_data <- generate_dynamics_data()
dynamics_results <- analyze_dynamics(dynamics_data)
dynamics_plot <- visualize_dynamics(dynamics_results)

# 4. Indicators
indicators_data <- generate_indicators_data()
indicators_plot <- visualize_indicators(indicators_data)

# Display alignment as example
print(alignment_plot)
```

## Interpreting Results

### Alignment Scores

- **0.80-1.00**: Strong alignment; shared vision well-established
- **0.60-0.79**: Moderate alignment; some areas need attention
- **Below 0.60**: Low alignment; significant discussion needed

### Cascade Scores

Higher scores indicate information and power successfully distributed across network degrees, suggesting sustainable community impact.

### Dynamics Scores

Scores reflect adherence to CBPR principles. Track changes over time to assess partnership development.

### Indicators

Contextualize traditional metrics within alignment, cascade, and dynamics scores for comprehensive impact assessment.

## Preparing Your Own Data

Each analysis function expects data in specific formats. Use the `generate_*_data()` functions as templates:

```{r data-structure}
# Examine expected structure
str(generate_alignment_data())
str(generate_cascade_data())
str(generate_dynamics_data())
str(generate_indicators_data())
```

## Next Steps

- See `?analyze_alignment` for detailed parameter descriptions
- Explore `?visualize_alignment` for customization options
- Review individual function documentation for each dimension
- Consult the CEnTR*IMPACT framework report for theoretical foundations

## References

Price, J. F. (2024). *CEnTR\*IMPACT: Community Engaged and Transformative Research – Inclusive Measurement of Projects & Community Transformation* (CUMU-Collaboratory Fellowship Report). Coalition of Urban and Metropolitan Universities. https://cumuonline.org/wp-content/uploads/2024-CUMU-Collaboratory-Fellowship-Report.pdf

Wallerstein, N., & Duran, B. (2010). Community-Based Participatory Research Contributions to Intervention Research: The Intersection of Science and Practice to Improve Health Equity. *American Journal of Public Health*, 100(S1), S40–S46. https://doi.org/10.2105/AJPH.2009.184036

Wallerstein, N., et al. (2020). Engage for Equity: A Long-Term Study of Community-Based Participatory Research and Community-Engaged Research Practices and Outcomes. *Health Education & Behavior*, 47(3), 380–390. https://doi.org/10.1177/1090198119897075
